\documentclass[12pt]{article}

%--- Packages ---%
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{authblk}
\usepackage{mathtools} % in preamble

\onehalfspacing
\bibliographystyle{aer}

%--- Theorem Environments ---%
\newtheorem{assumption}{Assumption}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}

%--- Title ---%
\title{\textbf{Love of Variety and Strategic Advertising:\\
               Evidence from Scanner Data}}

\author[1]{Tate Mason}
\affil[1]{John Munro Godfrey Sr. Department of Economics, University of Georgia}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
This paper develops a dynamic discrete choice model of consumer demand that incorporates a love of variety and strategic advertising by firms. The model is estimated using scanner data, and counterfactual analyses are conducted to evaluate the effects of different market structures on consumer welfare and market outcomes. The paper contributes to the literature on advertising and consumer behavior, by providing empirical evidence on the effects of strategic advertising in a dynamic demand setting. The findings of this paper have important implications for policymakers and firms, as they highlight the potential benefits and drawbacks of targeted advertising in markets when accounting for consumers' preference for variety.
\end{abstract}

\noindent\textbf{Keywords:} discrete choice, love of variety, habit formation,
                             structural estimation, advertising, demand estimation

\newpage

%============================================================%
\section{Introduction}
%============================================================%

In the baseline logit model, a consumer is perceived as having a fixed preference for each product, and the utility from consuming a product is independent of past consumption. Further, it is assumed that a new good is always a utility boosting event, by the error term's property of being identically and independently distributed. However, in many markets, consumers may have a love of variety, or a desire to consume different products over time. This can be due to factors such as boredom, satiation, or a preference for novelty. In such cases, the utility from consuming a product may depend on past consumption, and the introduction of a new good may not always lead to an increase in utility.

This paper develops a dynamic discrete choice model of consumer demand that incorporates a love of variety, and estimates the model using scanner data. The model allows for both forward-looking and myopic consumers, and includes a state variable that captures the mean consumption of each product. The model also incorporates strategic advertising by firms, which can influence consumer preferences and demand.

This idea is motivated by the idea that there are markets which see firms both acting as the host of the market as well as a participant in it. For example, in the grocery market, a firm like Kroger is both a retailer that hosts other firms' products, as well as a producer of its own goods. This may present a market power dynamic where the player with a dual role may have more access to consumer histories and preferences, allowing for more effective advertising via targeting. This paper seeks to understand the implications of this dynamic for consumer welfare and market outcomes. This will be accomplished by estimating the model using scanner data, and conducting counterfactual analyses to evaluate the effects of different market structures.

The primary contribution of this paper is wedding the love of variety and targeted advertising literatures, which have largely been separate, taking the other as given. Further, the firm side dynamics have been largely theoretical, and this paper seeks to provide empirical understanding of the effects of strategic advertising in a dynamic demand setting. Finally, the paper contributes to the broader literature on dynamic discrete choice models, by developing a model that incorporates both forward-looking and myopic consumers, and by providing an estimation strategy that can be applied to other settings.

%============================================================%
\section{Related Literature}
%============================================================%

    \subsection{Discrete Choice Demand Estimation}

    \subsection{Variety-Seeking and State Dependence}
    The variety-seeking and state dependence literature has been primarily focused on understanding the dynamics of consumer preferences and demand over time. From \cite{lattin_model_1987}, who developed an empirical structural model of state dependence in consumer choice, to \citet{erdem_decision-making_1996}, who iterated the model to allow for consumer learning. \cite{chintagunta_variety_1999} is the most closely related paper, as he develops a model of variety-seeking behavior which is both tractible and flexible, and uses Nielsen data to estimate the model. However, this paper only looks at demand, leaving me a niche to explore the interplay between variety-seeking and strategic advertising.


    \subsection{Advertising and Dynamic Demand}
    \citet{nelson_information_1970} is one of the earliest related papers, as he develops a model of advertising as a signal of product quality.

%============================================================%
\section{Model}
%============================================================%

    \subsection{Environment}
    There is a continuum of consumers $i\in\mathcal{I}$ who shop for $t\in\mathcal{T}$ periods. Each period $t$, they can make one purchase from a market. The market is made up of a continuum of firms $f\in\mathcal{F}$ in which one firm $f^*$ acts as both the market and a producer. $f^*$ will be the primary player, strategically advertising to consumer $i$ based on inferences about their type. $f^*$ will see all other $f\in\mathcal{F}$ as a competitor, denoted $\bar{f}$. The consumer is only concerned with maximizing their utility, while the firm wants to maximize profits.


    \subsection{Consumer's Problem}
    The consumer's problem is the following. In each period $t$, the consumer $i$ chooses a product $j$ from the menu of available products $J_t$. The consumer's utility from choosing product $j$ at time $t$ is given by:

    \begin{alignat*}{1}
        U_{ijt} &= \beta_i X_{jt} + \gamma_i \log(1 + (X_{jt} - \theta_{it})^2) 
                  + \alpha_i p_{jt}(1 - a_{it}) + \xi_j + \varepsilon_{ijt} \\
    \shortintertext{subject to}
        \theta_{it} &= \{f(X_{js})\}_{s < t} \\
        \gamma_i &\sim \mathcal{N}(0, 1)\\ \varepsilon_{ijt} &\sim \text{T1EV}(0,1)
    \end{alignat*}

    where $\beta_i$ is a consumer's utility from characteristic $X_{jt}$, $\gamma_i$ is their preference for variety, $\theta_{it}$ is a function of the consumers' past purchases. This could take the form of the mean characteristic of past purchases, the median characteristic, the norm, etc. $\alpha_i$ is a hetergeneous price sensitivity to $p_{jt}$, which is affected by some advertising $a_{it}$ in the form of coupons. $\xi_j$ is unobserved quality in product $j$, and $\varepsilon_{ijt}$ is the Gumbel shock which allows for a closed form logit.

    The choice probability of consumer $i$ choosing product $j$ in time $t$ is as follows:
    \begin{alignat*}{1}
      P_{ijt} = \frac{\exp{U_{ijt}}}{\sum_k\exp{U_{ikt}}}
    \end{alignat*}

    I am currently working under a specification in which the consumer is a static agent, but hope to expand to a forward-looking learning agent after the simple case is handled.

    \subsection{Firm's Problem}

    The firm $f^*$ chooses advertising $a_{it}$ to maximize profits. The firm's problem is as follows:
    \begin{alignat*}{1}
      V(a_{it} | \lambda_t) &= \max_{a_{it}} \{\pi_{jt} + \mu \mathbb{E}[V(a_{it+1} | \lambda_{t+1})]\} \\
      \pi_{jt} &= s(p_{jt}, a_{it})(p_{jt} - mc_{jt}) - c(a_{it}) \\
      \lambda_t &\sim \mathcal{N}(\frac{1}{N}\sum_i \theta_{it}, \sigma^2_\lambda)
    \end{alignat*}

    Here, $\pi_{jt}$ is the profit from advertising $a_{it}$, which depends on the market share $s(p_{jt}, a_{it})$, the price-cost margin $(p_{jt} - mc_{jt})$, and the cost of advertising $c(a_{it})$. The firm also has a discount factor $\mu$ for future profits, and faces uncertainty about the state variable $\lambda_t$, which is a function of the mean consumer history.

    Other firms $\bar{f}$ are not strategic and do not advertise, but they do compete in the market, affecting the consumer's choice probabilities and thus the profits of $f^*$. Their problem is as follows:
    \begin{alignat*}{1}
      \pi_{jt} &= s(p_{jt})(p_{jt} - mc_{jt})
    \end{alignat*}

    Most of the strategic dynamics will come from $f^*$, but the presence of $\bar{f}$ is important for identification and estimation.

    \subsection{Equilibrium}
    In equilibrium, consumers maximize utility via their consumption choices, and the firm maximized profits via advertising choices. The equilibrium concept is a Markov Perfect Equilibrium, in which the firm's advertising strategy is a function of the state variable $\lambda_t$, and the consumers' choice probabilities are consistent with their utility maximization given the advertising strategy. The equilibrium conditions can be expressed as follows:
    \begin{alignat*}{1}
      P_{ijt}^* &= \frac{\exp{U_{ijt}}}{\sum_k\exp{U_{ikt}}} \\
      a_{it}^* &= \arg\max_{a_{it}} \{\pi_{jt} + \mu \mathbb{E}[V(a_{it+1} | \lambda_{t+1})]\}
    \end{alignat*}

    where $P_{ijt}^*$ is the choice probability of consumer $i$ choosing the optimal product $j$ at time $t$, and $a_{it}^*$ is the optimal advertising strategy for firm $f^*$ given the state variable $\lambda_t$.

%============================================================%
\section{Estimation}
%============================================================%

    \subsection{Data}
    To estimate the model, I am in the process of acquiring Nielsen's homescan and marketscan data. This data combination will provide detailed information on consumer purchases, such as products bought, prices paid, and quantities purchased. Furhter, I am able to see deals/coupons which are used as well as any advertising in the form of feature displays. The data has a panel structure, which allows me to implement these dynamics. The data also includes information on product characteristics, which will be important for estimating the utility function. 

    \subsection{Identification of Random Coefficients}

    I plan to identify the random coefficients $\beta_i$, $\gamma_i$, and $\alpha_i$ using variation in product characteristics, prices, and advertising across consumers and over time. The presence of the strategic firm $f^*$ and its advertising will provide additional variation that can help identify the effects of advertising on consumer preferences and demand. I will also leverage the panel structure of the data to control for unobserved heterogeneity across consumers.

    Specifically, for the $\gamma_i$ coefficient, I will use variation in consumption over time. This includes mean number of prodcuts chosen, variance of products chosen, and number of times a conusmer switches between products. For the $\alpha_i$ coefficient, I will use variation in prices and income across consumers. For $\beta_i$, I will use variation in product characteristics and consumer demographics. The presence of the strategic firm $f^*$ and its advertising will provide additional variation that can help identify the effects of advertising on consumer preferences and demand. I will also leverage the panel structure of the data to control for unobserved heterogeneity across consumers.

    \subsection{GMM Objective and Moment Conditions}

      I will use a GMM estimation strategy to estimate the parameters of the model. The moment conditions will be derived from the first-order conditions of the consumer's utility maximization problem and the firm's profit maximization problem. Specifically, I will use the following moment conditions:
      \begin{alignat*}{1}
        \mathbb{E}[P_{ijt} - \frac{\exp{U_{ijt}}}{\sum_k\exp{U_{ikt}}}] &= 0 \\
        \mathbb{E}[a_{it} - \arg\max_{a_{it}} \{\pi_{jt} + \mu \mathbb{E}[V(a_{it+1} | \lambda_{t+1})]\}] &= 0
      \end{alignat*}

        These moment conditions ensure that the estimated parameters are consistent with the observed choice probabilities and advertising strategies in the data.

    \subsection{CCP Approach}

      To solve the dynamic programming problem of the firm, I will use a Conditional Choice Probability (CCP) approach. This approach allows me to estimate the value function and the optimal advertising strategy without having to solve the full dynamic programming problem. Instead, I will estimate the CCPs directly from the data, and then use these estimates to recover the value function and the optimal advertising strategy. This approach is computationally efficient and can handle the high-dimensional state space that arises from the consumer's past consumption history and the firm's strategic advertising decisions.

%============================================================%
\section{Monte Carlo Simulation}
%============================================================%

    \subsection{Design}
    To get a better understanding of the model, I conduct Monte Carlo simulations. I simulate data from the model under different parameter values for a singular consumer and no firm. This allows me to understand the dynamics of the model and the impact of the love of variety on choice probabilities. I simulate the model for 100 periods across 1000 simulations, and evaluate the effect of different levels of $\gamma_i$ on the choice probabilities. I also evaluate how $\gamma_i$ effects consumer welfare in the case of product entry.

    \subsection{Base Specification}

    \begin{equation}
      U_{ijt} = \beta_i X_{jt} + \varepsilon_{ijt}
    \end{equation}

    In this specification, I set $\gamma_i = 0$, which means that the consumer does not have a love of variety. This allows me to evaluate how the model behaves in the absence of variety-seeking behavior, and to establish a baseline for comparison with the other specifications. This is the specification which most closely resembles the standard logit model, and I find that the consumer has a flat consumption pattern, with no preference for variety.

    \input{../Output/Tables/tab_lov_0.tex}

    We see that almost all mass is on product 5, which has the highest characteristic value. This is consistent with the standard logit model, in which consumers always choose the product with the highest utility. The majority of the remaining mass in on product 4, which has the second highest characteristic value, a case which is created by the Gumbel error term. The consumer has no preference for variety, and thus does not switch between products over time.

    \subsection{Fixed Mean Specification}
    First, I use a fixed mean specification for $\theta_{it}$, in which the consumer's past consumption is summarized by a fixed mean of past product characteristics. This allows me to evaluate how the love of variety affects consumer choices and welfare when the consumer's preferences are influenced by a simple summary of their past consumption. I find a flat consumption pattern in which the consumer mostly picks the two polar products.

    \subsection{Prior Mean Specification}
    In this specification, I simulate a "history" for the consumer, and use the mean of that history to infer the initial $\theta_{i0}$. This allows me to evaluate how the love of variety affects consumer choices and welfare when the consumer's preferences are influenced by a more complex summary of their past consumption. I find that the consumer has a more varied consumption pattern, and is more likely to switch between products for the first 20 periods before settling into a pattern of mostly picking the two polar products.

    \begin{table}[h]
    \centering
    \caption{Conditional Choice Probabilities by $\gamma$}
    \begin{tabular}{lccc}
    \hline\hline
    & $\gamma = 6$ & $\gamma = 9$ & $\gamma = 12$ \\
    \hline
    Product 1       & 0.3313  & 0.3971  & 0.4234  \\
    Product 2       & 0.0551  & 0.0081  & 0.0009  \\
    Product 3       & 0.0027  & 0.0000  & 0.0000  \\
    Product 4       & 0.0023  & 0.0002  & 0.0000  \\
    Product 5       & 0.6086  & 0.5946  & 0.5756  \\
    \hline
    Inclusive Value & 17.4522 & 22.4506 & 27.5139 \\
    IV at $t=50$    & 16.3861 & 21.5165 & 27.6856 \\
    \hline\hline
    \end{tabular}
    \end{table}

    As $\gamma$ increases, the consumer's preference for variety increases, and thus they are more likely to switch between products over time. This is reflected in the choice probabilities, which show that as $\gamma$ increases, the consumer is more likely to choose product 1, which has the lowest characteristic value, and less likely to choose product 5, which has the highest characteristic value. The consumer's love of variety leads them to seek out different products over time, even if those products have lower utility than their previous choices. We also see that the inclusive value increases as $\gamma$ increases, which indicates that the consumer's overall utility from the available products increases as their preference for variety increases. This is consistent with the idea that a love of variety can lead to higher consumer welfare, as it allows consumers to derive more utility from a wider range of products.


%============================================================%
\section{Counterfactuals}
%============================================================%

    \subsection{Counterfactual 1}

    In my first idea, I want to evaluate the effects of a restriction on the strategic firm's ability to target based on competitors' purchase information. This could be motivated by a policy that restricts the use of data to consolidate market power. In this counterfactual, I will simulate the model under the assumption that strategic firm $f^*$ now maximizes like the non-strategic firms $\bar{f}$, and does not have access to consumer history or preferences. This will allow me to evaluate the effects of strategic advertising on consumer welfare and market outcomes, and to understand the implications of such a policy.

    \subsection{Counterfactual 2}

    Secondly, I am interested in pivoting to a more consumer welfare focused counterfactual, in which I evaluate the effects of product entry on consumer welfare under different levels of $\gamma$. This will allow me to understand how a love of variety affects consumer welfare in the face of new product entry, and to evaluate the potential benefits and drawbacks of a love of variety for consumers in dynamic markets. Specifically, I want to understand how consumer welfare is impacted by both perfect and imperfect information about the distribution of $\gamma$ when introducting a new product.


%============================================================%
\section{Conclusion}
%============================================================%

In closing, this paper develops a dynamic discrete choice model of consumer demand that incorporates a love of variety and strategic advertising by firms. The model is estimated using scanner data, and counterfactual analyses are conducted to evaluate the effects of different market structures on consumer welfare and market outcomes. The paper contributes to the literature on  advertising and consumer behavior, by providing empirical evidence on the effects of strategic advertising in a dynamic demand setting. The findings of this paper have important implications for policymakers and firms, as they highlight the potential benefits and drawbacks of targeted advertising in markets when accounting for consumers' preference for variety. Future research could extend this model to incorporate additional features, such as consumer learning, network effects, or multi-product firms, to further understand the dynamics of consumer demand and firm behavior in these markets.

%============================================================%
\newpage

\bibliography{lov_lit.bib}

\end{document}
